---
title: |
  **Forecasting Government Expenditure in Ghana Using Autoregressive Integrated Moving Average with Exogenous Variables (ARIMAX)**
author: "Gershon Tetteh Amanor"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 4
    number_sections: true
    theme: flatly
    highlight: tango
    df_print: kable
    fig_width: 10
    fig_height: 5.5
    code_folding: show
  pdf_document:
    toc: true
    toc_depth: '4'
subtitle: null
editor_options: 
  markdown: 
    wrap: 72
---

```{r global-setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  comment = "#>",
  fig.align = "center"
)
options(scipen = 999)
```

------------------------------------------------------------------------

> **Abstract:** This study develops and evaluates an Autoregressive
> Integrated Moving Average with Exogenous Variables (ARIMAX) model for
> forecasting Ghana's government expenditure over the period 1984–2024.
> The dependent variable is log-transformed total government expenditure
> (USD equivalent), and three exogenous variables are examined —
> population, log government revenue, and log GDP — both jointly and in
> isolation. The ARIMAX framework is compared against a univariate ARIMA
> baseline using information criteria (AIC, BIC) and standard accuracy
> metrics (RMSE, MAE, MAPE, MASE). Results demonstrate that
> incorporating macroeconomic exogenous variables into the time series
> framework meaningfully improves forecasting performance. Policy
> implications for Ghana's medium-term fiscal planning are discussed.

------------------------------------------------------------------------

# Chapter One: Introduction {.tabset}

## Background of Study

Government expenditure is a critical pillar of Ghana's development,
financing key sectors such as education, healthcare, infrastructure, and
social protection that contribute directly to productivity growth, human
capital development, and long-term economic performance (World Bank,
2023). In the majority of developing countries, government is seen as an
instrument of change, and hence the size of government expenditure
reveals the magnitude of government involvement in the economy.

The shift in the role of government — from traditional functions such as
provision of security, administration, and law and order, to direct
intervention in income-generating activities like capital investment and
distributive roles like subsidies and transfers — has significantly
expanded the scope of governments in many countries across the globe.

Government expenditure is defined as the total spending by the
government on recurrent items and capital projects, primarily funded
through budgetary appropriations and often involving significant
allocations for personnel costs (Leonard Onyiruba, 2016). In Ghana,
government expenditure includes various components such as total public
spending on purchase of goods and services, compensation of employees,
interest payments, grants to other government units, non-financial
assets, and capital expenditure. As defined by the International
Monetary Fund (IMF), general government expenditure consists of total
expense and the net acquisition of nonfinancial assets (Statista, 2025).

**Ghana's fiscal exposure** is characterised by its status as a
commodity-dependent economy. Changes in world prices of cocoa, gold, and
oil directly affect export earnings, tax revenues, and the overall
fiscal balance. When commodity prices rise, revenue and spending
pressures typically rise; when prices fall, fiscal stress often triggers
spending cuts or arrears. Ghana's public spending has fluctuated
significantly due to changes in political priorities, inflationary
pressures, and debt servicing obligations — inconsistencies that make it
difficult for policymakers to predict expenditure accurately, often
resulting in fiscal imbalances and budget deficits.

These trends underscore a critical need for reliable forecasting tools
that can help policymakers anticipate future expenditure paths, identify
potential fiscal risk, and make informed, data-driven decisions that
ensure long-term fiscal sustainability.

Traditional estimation methods have limitations in handling dynamic
economic data; hence time-series models such as the **ARIMAX**
(Autoregressive Integrated Moving Average with Exogenous Variables)
offer a more reliable approach. In many real situations, the assumption
that only past values influence future outcomes is too limited because
external factors often play a significant role in influencing the
direction and behaviour of a time series. This is where ARIMAX becomes
particularly valuable — by integrating exogenous variables directly into
the forecasting process, allowing for more realistic and accurate
predictions.

## Problem Statement

Government expenditure in Ghana faces several significant challenges and
limitations, primarily driven by high public debt, a narrow revenue
base, and weaknesses in public financial management (PFM) systems.
Despite fiscal reforms, Ghana continues to face budget overruns,
expenditure volatility, and debt-related pressures. The May 2023 IMF
Extended Credit Facility (ECF) arrangement was triggered by severe
fiscal and debt vulnerabilities following external shocks and policy
slippages (Ministry of Finance, Ghana).

Awariefe & Ogumeyo (2023) found that traditional regression (MLR)
underperforms when predicting government expenditure because it cannot
handle complex, non-linear macro dynamics. Djakou & Jiang (2023) found
that an ARIMA(1,1,1) model assumes a linear trend and does not fully
capture external macro shocks such as volatility in budget performance.
While several studies have applied ARIMA-type models in Ghana for
forecasting inflation, GDP, or tax revenue, there is limited evidence of
similar applications for **government expenditure forecasting using the
ARIMAX framework**.

This methodological gap reduces policymakers' ability to anticipate the
implications of macroeconomic shocks and IMF fiscal regimes. This study
addresses the gap by applying ARIMAX with population, revenue, and GDP
as exogenous variables to deliver more reliable fiscal forecasts for
Ghana.

## Objectives of the Study

**General Objective:** To develop an ARIMAX-based forecasting model for
Ghana's government expenditure, using population, government revenue,
and GDP as exogenous variables.

**Specific Objectives:**

1.  To examine historical trends and volatility in Ghana's government
    expenditure (1983–2024).
2.  To assess the influence of macroeconomic variables (population,
    revenue, GDP) on government expenditure.
3.  To estimate optimal ARIMAX specifications and generate out-of-sample
    forecasts.
4.  To evaluate whether ARIMAX with exogenous variables improves
    forecasting accuracy relative to a univariate ARIMA baseline.

## Research Questions

1.  What are the dominant trends and fluctuations in Ghana's government
    expenditure over time?
2.  How do population growth, government revenue, and GDP movements
    affect government expenditure?
3.  Does an ARIMAX model with macroeconomic exogenous variables improve
    forecasting accuracy relative to ARIMA?
4.  Which combination of exogenous variables yields the most
    parsimonious and accurate ARIMAX specification?

## Significance of the Study

This study is significant on multiple fronts:

-   **Policy relevance:** Improves expenditure forecasts that help the
    Ministry of Finance and Parliament set realistic budget ceilings,
    reduce expenditure arrears, and better manage fiscal risks.
-   **Methodological contribution:** Demonstrates the effectiveness of
    parsimonious ARIMAX modelling in a developing, commodity-dependent
    economy, contributing to the applied time-series econometrics
    literature.
-   **Institutional utility:** Provides a replicable, transparent, and
    data-driven analytical framework that can be adopted by the Bank of
    Ghana, the National Development Planning Commission, and development
    partners for medium-term expenditure planning.

------------------------------------------------------------------------

# Chapter Two: Literature Review {.tabset}

## Theoretical Foundations

The theoretical framework for this study draws on several interrelated
strands of economic and statistical theory:

**Box-Jenkins ARIMA Methodology**\
The theoretical foundation for ARIMA (Autoregressive Integrated Moving
Average) modelling is based on the Box-Jenkins methodology developed by
George Box and Gwilym Jenkins (1970), which establishes a systematic
process for identification, estimation, and diagnostic checking of time
series models. ARIMAX extends this framework by incorporating exogenous
regressors (transfer functions), allowing the model to capture external
influences on the primary series.

**Keynesian Fiscal Theory**\
John Maynard Keynes (1936) posits that government spending is a critical
component of aggregate demand and a powerful tool for economic
stabilisation. This perspective reinforces the importance of accurately
forecasting expenditure to guide effective fiscal planning and ensure
macroeconomic stability. According to Keynesian theory, an increase in
government expenditure leads to an increase in aggregate demand,
production, and ultimately GDP — making GDP a natural exogenous
candidate in the ARIMAX framework.

**Wagner's Law**\
Named after German economist Adolph Wagner, this hypothesis posits that
as an economy develops over time, government expenditure grows more than
proportionally relative to national income. Wagner's Law implies a
long-run upward drift in public spending as economies grow, providing
the theoretical basis for including GDP and population as exogenous
drivers of expenditure.

**Peacock-Wiseman Displacement Hypothesis**\
Peacock and Wiseman (1961) argue that public expenditure does not
increase at a steady rate, but rather jumps in response to major
economic or social disturbances (wars, crises, pandemics). Once
elevated, spending rarely returns to its pre-shock level — a
"displacement" effect. This theoretical lens is important for
understanding Ghana's expenditure jumps around periods of economic
crisis.

**Commodity-Price Fiscal Linkage**\
In commodity-exporting economies, global price shocks influence tax
receipts, borrowing costs, and expenditure choices. IMF working paper
evidence (Spatafora & Samake, 2012) shows both revenue and expenditure
respond significantly to commodity price movements, making fiscal policy
highly procyclical in such settings.

## Empirical Literature

A growing body of literature confirms that ARIMAX models consistently
outperform baseline ARIMA models when relevant exogenous information is
available:

-   **Jiayi Yang (2025):** Applied ARIMA, ARIMAX, VAR, and GLM to
    forecast China's university admission rates (1977–2024). The ARIMAX
    model, incorporating GDP and newborn population as exogenous
    variables, produced the lowest forecast errors and provided deeper
    insights into how macroeconomic factors drive admission rates.

-   **Fairuz Azizah et al. (2025):** Modelled Indonesia's Human
    Development Index using ARIMAX with inflation and infant mortality
    as exogenous variables. Results confirmed that including these
    external drivers significantly improved estimation accuracy over the
    pure ARIMA specification.

-   **Coker (2025):** Compared ARIMA, ARIMAX, and VAR for short-term
    inflation forecasting in Sierra Leone. The ARIMAX model
    incorporating the exchange rate as an exogenous variable
    significantly outperformed the alternatives, underscoring the
    importance of external drivers in small, open economies.

-   **Abdulazeez (2021):** Demonstrated in a Nigeria GDP forecasting
    study that ARIMAX with agriculture, industry, and services sector
    shares consistently outperformed ARIMA across all error metrics
    (RMSE: 899 vs 1,662; MSE: 568 vs 1,161).

-   **Ajibode & Adeboye:** Applied ARIMA and ARIMAX to COVID-19
    mortality modelling in Nigeria. ARIMAX with confirmed cases as an
    exogenous variable produced the lowest MAE (5.98) and RMSE (8.54),
    outperforming the pure ARIMA specification.

-   **Michael Safo Ofori et al.:** Forecasting VAT revenue in Ghana
    using ARIMA with intervention, demonstrating the utility of
    time-series methods for Ghanaian fiscal data and validating the
    approach adopted in the present study.

-   **Boachie et al. (2022):** Examined Ghana's public health spending
    (1980–2014) using 3SLS estimators, finding that IMF programme
    conditionalities slightly constrained public health spending,
    supporting the inclusion of IMF regime indicators in fiscal
    forecasting models.

## Chapter Summary

The reviewed literature consistently highlights that: (1) ARIMA models
are a well-established baseline for univariate fiscal forecasting; (2)
the incorporation of exogenous variables through the ARIMAX framework
reliably improves forecasting accuracy; (3) macro-fiscal variables such
as GDP, revenue, and population are the most empirically validated
exogenous drivers; and (4) there is a notable gap in the application of
ARIMAX to Ghana's government expenditure forecasting — the gap this
study addresses.

------------------------------------------------------------------------

# Chapter Three: Methodology

## Research Design

A **quantitative, predictive time-series design** is adopted, consistent
with the Box-Jenkins tradition. The study employs secondary annual data
covering the period **1983–2024** (42 observations), drawing on official
government fiscal records and national accounts. The research is both
explanatory (examining the influence of macroeconomic exogenous
variables on expenditure) and predictive (generating a 10-year
out-of-sample forecast).

## Data Sources and Variables

```{r variable-table, echo=FALSE}
var_table <- data.frame(
  Variable       = c("GE (Y)", "POPULATION (X1)", "REVENUE (X2)", "GDP (X3)"),
  Description    = c("Total Government Expenditure (Net Lending)",
                     "Ghana's Annual Population",
                     "Total Government Revenue (Million GHS → USD)",
                     "Nominal GDP (USD Billion → USD)"),
  Unit           = c("Million GHS → log(USD)", "Persons", "log(USD)", "log(USD)"),
  Source         = c("Bank of Ghana — Gov. Fiscal Operations",
                     "World Bank / Ghana Statistical Service",
                     "Bank of Ghana — Gov. Fiscal Operations",
                     "World Bank National Accounts")
)
knitr::kable(var_table, caption = "Table 1: Variable Description and Sources", align = "l")
```

The dataset spans **1983 to 2024**, with the following coverage:

-   **Government Expenditure:** Annual total expenditure and net lending
    (millions of Ghana Cedis), sourced from the Bank of Ghana's
    Government Fiscal Operations report.
-   **Population:** Annual population figures used as a demographic
    driver of public expenditure demand.
-   **Revenue:** Annual government revenue (millions of Ghana Cedis),
    converted to USD and log-transformed.
-   **GDP:** Nominal GDP in USD billions, converted to USD and
    log-transformed.

All currency-denominated variables (expenditure, revenue) are
**converted from Ghana Cedis to US Dollars** using the contemporaneous
annual average exchange rate. This conversion normalises the data
against the extreme nominal exchange-rate depreciation that Ghana
experienced over the study period, particularly post-2014. **Log
transformations** are then applied to all monetary variables to
stabilise variance and linearise exponential growth trends.

## Analytical Framework: The ARIMAX Model

### ARIMA Components

The ARIMA(p, d, q) model combines three core components:

-   **AR(p) — Autoregressive component:** Models the relationship
    between the current observation and $p$ lagged values:
    $$X_t = c + \phi_1 X_{t-1} + \phi_2 X_{t-2} + \ldots + \phi_p X_{t-p} + \epsilon_t$$

-   **I(d) — Integration (Differencing):** The series is differenced $d$
    times to achieve stationarity, removing trend and unit-root
    behaviour.

-   **MA(q) — Moving Average component:** Models the relationship
    between the current value and $q$ lagged forecast errors:
    $$X_t = c + \epsilon_t + \theta_1 \epsilon_{t-1} + \theta_2 \epsilon_{t-2} + \ldots + \theta_q \epsilon_{t-q}$$

Combined:
$$X_t = c + \sum_{i=1}^{p}\phi_i X_{t-i} + \sum_{j=1}^{q}\theta_j \epsilon_{t-j} + \epsilon_t$$

### ARIMAX Extension

The ARIMAX model augments the ARIMA framework with exogenous regressors
$X_k$:

$$Y_t = \mu + \sum_{i=1}^{p}\phi_i Y_{t-i} + \sum_{j=1}^{q}\theta_j \epsilon_{t-j} + \sum_{k}\beta_k X_{k,t} + \epsilon_t$$

where: - $Y_t$ = Government Expenditure (log USD) at time $t$ -
$X_{k,t}$ = exogenous variables (population, log revenue, log GDP) -
$\beta_k$ = coefficients for the exogenous variables -
$\epsilon_t \sim WN(0, \sigma^2)$

### Model Selection Criteria

| Criterion | Formula                        | Preference      |
|-----------|--------------------------------|-----------------|
| AIC       | $-2\log(L) + 2k$               | Lower is better |
| BIC       | $-2\log(L) + k\log(n)$         | Lower is better |
| RMSE      | $\sqrt{\frac{1}{n}\sum e_t^2}$ | Lower is better |
| MAE       | $\frac{1}{n}\sum|e_t|$         | Lower is better |
| MAPE      | $\frac{100}{n}\sum|e_t/Y_t|$   | Lower is better |
| MASE      | Scaled by naïve forecast MAE   | \< 1 is better  |

### Stationarity Tests

**Augmented Dickey-Fuller (ADF) Test:**
$$\Delta y_t = \alpha + \beta t + \rho y_{t-1} + \sum_{i=1}^{k}\phi_i \Delta y_{t-i} + \epsilon_t$$

-   $H_0$: Unit root present (non-stationary)
-   $H_1$: Series is stationary
-   Reject $H_0$ if p-value \< 0.05

------------------------------------------------------------------------

# Chapter Four: Data Analysis and Results

## Setup — Libraries and Global Options

All required R packages are loaded here. These cover time series
modelling (`forecast`, `tseries`), multicollinearity diagnostics
(`car`), data import (`readxl`), and data manipulation/visualisation
(`tidyverse`).

```{r ch4-libraries}
# Time series modelling and forecasting toolkit
library(forecast)

# Augmented Dickey-Fuller and KPSS unit root tests
library(tseries)

# Variance Inflation Factor for multicollinearity diagnostics
library(car)

# Reading Excel data files
library(readxl)

# Data manipulation and visualisation (dplyr, ggplot2, tidyr)
library(tidyverse)
```

------------------------------------------------------------------------

## Data Loading and Preprocessing

### Loading the Dataset

```{r ch4-load}
# Load the annual fiscal dataset (1983–2024)
data_raw <- read_excel("Data2.xlsx")

# Preview the raw data structure
glimpse(data_raw)
```

### Currency Conversion and Log Transformation

Government expenditure and revenue are denominated in **millions of
Ghana Cedis (GHS)**. To control for the severe nominal depreciation of
the cedi over the study period — from roughly 2.7 GHS/USD in 1984 to
over 12 GHS/USD by the early 2020s — all monetary variables are
**converted to US Dollars** using the contemporaneous annual exchange
rate.

Log transformations are applied to all monetary variables to: 1.
Stabilise the exponential growth in variance (heteroscedasticity
correction) 2. Linearise the long-run upward trend, making the series
more amenable to ARIMA modelling 3. Enable percentage-change
interpretation of first differences

```{r ch4-transform}
data <- data_raw |>
  mutate(
    # Convert GHS expenditure (millions) to USD
    GOV_EXP_DOLLARS = EXPENDITURE * 1000000 / EXCHANGE_RATE,

    # Convert GHS revenue (millions) to USD
    REVEVENUE = `REVENUE MILLION OF  CEDIS` * 1000000 / EXCHANGE_RATE,

    # Convert GDP (USD billions) to USD for unit consistency
    GDP = `GDP USD BILLION` * 1000000000,

    # Apply natural log transformation to all monetary variables
    # Rationale: log stabilises exponential variance growth and linearises trend
    LOG_EXPENDITURE = log(GOV_EXP_DOLLARS),
    LOG_REVENUE     = log(REVEVENUE),
    LOG_GDP         = log(GDP)
  ) |>
  # Remove original pre-transformation columns and the election dummy
  # (election effects are not part of this specification)
  select(-c(EXPENDITURE, `REVENUE MILLION OF  CEDIS`, `GDP USD BILLION`,
            EXCHANGE_RATE, ELECTION)) |>
  # Remove rows where division by zero produced Inf values
  filter(GOV_EXP_DOLLARS != Inf)

# Show the processed dataset structure
glimpse(data)
```

### Processed Data Summary

```{r ch4-summary}
summary(data[, c("GOV_EXP_DOLLARS", "LOG_EXPENDITURE",
                 "LOG_REVENUE", "LOG_GDP", "POPULATION")])
```

### Defining Time Series Objects

```{r ch4-ts-define}
# Dependent variable: log government expenditure as annual time series
# Starting from 1984 (first year with valid exchange rate data)
Y <- ts(data$LOG_EXPENDITURE, start = 1984, frequency = 1)

# Exogenous variable matrix used in ARIMAX models
xreg_vars <- data[, c("POPULATION", "LOG_REVENUE", "LOG_GDP")]

# Forecast horizon: 10 years beyond the end of the estimation sample
h_period <- 10

cat("Dependent variable (Y): Log Government Expenditure\n")
cat("Sample period:", start(Y)[1], "to", end(Y)[1], "\n")
cat("Observations:", length(Y), "\n")
cat("Forecast horizon:", h_period, "years\n")
```

------------------------------------------------------------------------

## Descriptive Analysis and Trend Exploration

### Historical Trend of Government Expenditure

Before formal modelling, it is important to visually inspect Ghana's
government expenditure history. The time series plot below reveals the
character of the data: the direction of the trend, any structural
breaks, and the general growth trajectory.

```{r ch4-descriptive-plot, fig.cap="Figure 1: Ghana's Log Government Expenditure in USD (1984–2024)", fig.height=6}
par(mfrow = c(2, 1), mar = c(4, 5, 3, 2))

# Plot 1: Raw USD expenditure
plot(ts(data$GOV_EXP_DOLLARS / 1e9, start = 1984),
     type = "o", col = "steelblue", lwd = 2,
     main = "Ghana Total Government Expenditure (USD Billion)",
     ylab = "USD Billion", xlab = "Year")
grid(col = "lightgray", lty = "dotted")

# Plot 2: Log-transformed expenditure (used as Y in models)
plot(Y,
     type = "o", col = "darkgreen", lwd = 2,
     main = "Log Government Expenditure (used as dependent variable Y)",
     ylab = "Log(USD)", xlab = "Year")
grid(col = "lightgray", lty = "dotted")
```

> **Observation:** The raw expenditure series exhibits exponential
> growth — characteristic of a nominal fiscal variable in a developing
> economy undergoing both real growth and significant price-level
> increases. The log transformation converts this to a near-linear
> upward trend, more appropriate for ARIMA-based modelling. Notable
> inflections (post-2014 cedi depreciation, post-2020 COVID-19 fiscal
> expansion) are visible in both plots.

### Exogenous Variables Over Time

```{r ch4-exog-plots, fig.cap="Figure 2: Exogenous Variables — Population, Log Revenue, Log GDP (1984–2024)", fig.height=7}
par(mfrow = c(3, 1), mar = c(3, 5, 2, 2))

plot(ts(data$POPULATION / 1e6, start = 1984),
     type = "o", col = "darkorange", lwd = 1.5,
     main = "Ghana Population (Millions)", ylab = "Millions", xlab = "")
grid(col = "lightgray", lty = "dotted")

plot(ts(data$LOG_REVENUE, start = 1984),
     type = "o", col = "purple", lwd = 1.5,
     main = "Log Government Revenue (USD)", ylab = "Log(USD)", xlab = "")
grid(col = "lightgray", lty = "dotted")

plot(ts(data$LOG_GDP, start = 1984),
     type = "o", col = "firebrick", lwd = 1.5,
     main = "Log GDP (USD)", ylab = "Log(USD)", xlab = "Year")
grid(col = "lightgray", lty = "dotted")
```

> **Observation:** All three exogenous variables exhibit persistent
> upward trends over the sample period. Population growth is smooth and
> demographic in nature. Revenue and GDP share a similar trajectory to
> expenditure, reflecting the co-movement between fiscal variables and
> macroeconomic performance over Ghana's development path. This
> co-movement is precisely the signal that ARIMAX is designed to
> exploit.

------------------------------------------------------------------------

## Multicollinearity Check (VIF)

Before including multiple exogenous regressors in the ARIMAX model,
potential **multicollinearity** is assessed using the **Variance
Inflation Factor (VIF)**. Multicollinearity arises when two or more
predictors are highly linearly correlated, inflating coefficient
standard errors and undermining interpretability.

**Decision rule:** - VIF \< 5 → Acceptable - VIF 5–10 → Moderate
concern; monitor - VIF \> 10 → Severe multicollinearity → consider
dimensionality reduction or variable exclusion

Given that population, revenue, and GDP are all trending upward over the
same period, some degree of multicollinearity is expected. This
motivates the individual ARIMAX models (one exogenous variable at a
time) fitted in Section 4.7–4.9.

```{r ch4-vif}
# Fit auxiliary OLS regression solely to compute VIF values
# (Not used for forecasting; diagnostic purpose only)
vif_model <- lm(Y ~ POPULATION + LOG_REVENUE + LOG_GDP, data = data)

cat("=== Variance Inflation Factors (VIF) ===\n")
print(vif(vif_model))

cat("\nNote: VIF > 10 indicates severe multicollinearity.\n")
cat("VIF > 5 suggests moderate collinearity that may inflate standard errors.\n")
```

> **Interpretation:** High VIF values among `LOG_REVENUE`, `LOG_GDP`,
> and `POPULATION` are plausible given that all three grow in tandem
> with the Ghanaian economy over time. The joint ARIMAX model (Section
> 4.6) is included for completeness, but the individual models (Sections
> 4.7–4.9) provide cleaner, more interpretable estimates by avoiding
> multicollinear combinations.

------------------------------------------------------------------------

## 4.5 Stationarity Testing and Order of Integration

### ADF Test on the Level Series

The ADF test evaluates whether the log expenditure series $Y_t$ contains
a unit root (is non-stationary):

-   **H₀:** Unit root present → series is non-stationary\
-   **H₁:** No unit root → series is stationary\
-   **Reject H₀** if p-value \< 0.05

```{r ch4-adf-level}
cat("=== ADF Test: Log Government Expenditure (Level) ===\n")
adf_level <- adf.test(Y)
print(adf_level)

if (adf_level$p.value > 0.05) {
  cat("\n>> Decision: Fail to reject H0 (p =", round(adf_level$p.value, 4),
      ") — Series is NON-STATIONARY. Differencing required.\n")
} else {
  cat("\n>> Decision: Reject H0 (p =", round(adf_level$p.value, 4),
      ") — Series is STATIONARY. No differencing needed.\n")
}
```

### Second-Order Differencing

If the level series is non-stationary, differencing is applied to remove
the trend and induce stationarity. Second-order differencing (d = 2)
corresponds to the integration order I(2), which is common for fiscal
series with strong exponential growth.

```{r ch4-adf-diff}
# Apply second-order differencing
log_y_diff <- diff(Y, differences = 2)

cat("=== ADF Test: Second-Order Differenced Series ===\n")
adf_diff <- adf.test(log_y_diff)
print(adf_diff)

if (adf_diff$p.value < 0.05) {
  cat("\n>> Decision: Reject H0 (p =", round(adf_diff$p.value, 4),
      ") — Differenced series is STATIONARY. d = 2 is appropriate.\n")
} else {
  cat("\n>> Decision: Fail to reject H0 — further differencing may be required.\n")
}
```

### Visual Comparison: Level vs Differenced Series

```{r ch4-stationarity-plots, fig.cap="Figure 3: Level and Second-Differenced Log Government Expenditure"}
par(mfrow = c(2, 1), mar = c(4, 5, 3, 2))

plot(Y, type = "o", col = "steelblue", lwd = 1.5,
     main = "Level Series: Log Government Expenditure",
     ylab = "Log(USD)", xlab = "")
abline(lm(Y ~ time(Y)), col = "red", lty = 2, lwd = 2)
legend("topleft", legend = "OLS Trend Line", col = "red",
       lty = 2, bty = "n", cex = 0.85)
grid(col = "lightgray", lty = "dotted")

plot(log_y_diff, type = "o", col = "firebrick", lwd = 1.5,
     main = "Second-Order Differenced Series (d = 2)",
     ylab = "Differenced Log(USD)", xlab = "Year")
abline(h = 0, col = "black", lty = 2)
grid(col = "lightgray", lty = "dotted")
```

> **Interpretation:** The level series (top panel) shows a clear upward
> trend with a fitted OLS trend line, confirming non-stationarity. After
> second-order differencing (bottom panel), the series fluctuates around
> zero without a systematic drift, indicating that the trend has been
> successfully removed. The ADF test result on the differenced series
> confirms stationarity, and d = 2 is adopted in the ARIMA/ARIMAX
> specifications.

------------------------------------------------------------------------

## ACF and PACF Analysis — Identifying Model Orders

The **Autocorrelation Function (ACF)** and **Partial Autocorrelation
Function (PACF)** are the primary tools for identifying the
autoregressive order (p) and moving average order (q) in the Box-Jenkins
framework:

| Pattern | Implication |
|------------------------------|------------------------------------------|
| PACF: spike at lag $k$, then cuts off | AR($k$) component |
| ACF: spike at lag $k$, then cuts off | MA($k$) component |
| Both decay slowly | Mixed ARMA; use `auto.arima` |
| No significant spikes in differenced series | Differencing sufficient; d is appropriate |

```{r ch4-acf-level, fig.cap="Figure 4: ACF and PACF of Level Series (Log Government Expenditure)"}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
Acf(Y,  main = "ACF — Level Series",  lag.max = 20); grid()
Pacf(Y, main = "PACF — Level Series", lag.max = 20); grid()
```

```{r ch4-acf-diff, fig.cap="Figure 5: ACF and PACF of Second-Differenced Series"}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
Acf(log_y_diff,  main = "ACF — Differenced Series",  lag.max = 20); grid()
Pacf(log_y_diff, main = "PACF — Differenced Series", lag.max = 20); grid()
```

> **Interpretation:** The level series ACF decays slowly (consistent
> with non-stationarity). In the differenced series, spikes that fall
> within the 95% confidence bands (blue dashed lines at ±1.96/√n)
> suggest that differencing has removed most autocorrelation. Remaining
> significant spikes, if any, guide the selection of p and q.
> `auto.arima()` in subsequent sections performs a systematic
> AIC/BIC-guided search to confirm the optimal orders.

------------------------------------------------------------------------

## Defining Forecast Variables

Future values of the exogenous variables must be specified to generate
ARIMAX out-of-sample forecasts. A **mean-projection approach** is used
here, setting each exogenous variable to its in-sample mean across the
10-year horizon. This is a conservative, assumption-neutral approach.

> In practice, scenario-based projections (e.g., IMF WEO GDP
> projections, UN population forecasts) would be preferred for policy
> applications. The mean projection provides a baseline forecast that
> captures the average historical influence of each exogenous variable.

```{r ch4-future-xreg}
# Mean-based projection of exogenous variables over 10-year horizon
future_population <- rep(mean(data$POPULATION), h_period)
future_revenue    <- rep(mean(data$LOG_REVENUE), h_period)
future_gdp        <- rep(mean(data$LOG_GDP),     h_period)

# Assemble as a matrix with column names matching the estimation xreg
future_xreg <- cbind(future_population, future_revenue, future_gdp)
colnames(future_xreg) <- colnames(xreg_vars)

cat("Future exogenous variable projections (mean of in-sample values):\n")
cat("Population (mean):", round(mean(data$POPULATION), 0), "\n")
cat("Log Revenue (mean):", round(mean(data$LOG_REVENUE), 4), "\n")
cat("Log GDP (mean):", round(mean(data$LOG_GDP), 4), "\n")
```

------------------------------------------------------------------------

## Model 1: ARIMA Baseline

The univariate ARIMA model serves as the benchmark against which all
ARIMAX specifications are evaluated. It explains Ghana's log expenditure
solely from its own historical dynamics — past values and lagged shocks
— with no external information.

### Model Estimation

`auto.arima()` with `stepwise = FALSE` and `approximation = FALSE`
performs an exhaustive search across candidate (p, d, q) combinations,
selecting the specification with the lowest AIC. Maximum Likelihood
Estimation (MLE) is used for parameter estimation, providing efficient
and asymptotically unbiased estimates.

```{r ch4-arima-fit}
cat("=== Fitting ARIMA Baseline Model ===\n")
cat("Method: Maximum Likelihood Estimation (MLE)\n")
cat("Selection: AIC-minimising exhaustive search\n\n")

arima_fit <- auto.arima(
  Y,
  stepwise      = FALSE,   # Exhaustive, not stepwise, search
  approximation = FALSE,   # Exact (not approximate) AIC values
  method        = "ML"     # Maximum Likelihood Estimation
)

summary(arima_fit)
```

### Residual Diagnostics

A valid ARIMA model should produce **white noise residuals** — residuals
that are uncorrelated, approximately normally distributed, and have
constant variance. Any remaining structure in the residuals indicates
model misspecification.

```{r ch4-arima-diag, fig.cap="Figure 6: ARIMA Baseline — Residual Diagnostics"}
checkresiduals(arima_fit)
```

### Ljung-Box Test for Residual Autocorrelation

The Ljung-Box test formally tests whether the first 10 residual
autocorrelations are jointly zero:

-   **H₀:** Residuals are uncorrelated (white noise) — model is
    adequate\
-   **H₁:** Significant autocorrelation remains — model may be
    misspecified\
-   **Rule:** p-value \> 0.05 → fail to reject H₀ → residuals are clean

```{r ch4-arima-lb}
lb_arima <- Box.test(residuals(arima_fit), lag = 10, type = "Ljung-Box")
cat("=== Ljung-Box Test (lag = 10) — ARIMA Baseline ===\n")
print(lb_arima)

if (lb_arima$p.value > 0.05) {
  cat("\n>> Decision: p =", round(lb_arima$p.value, 4),
      "> 0.05 — Fail to reject H0. Residuals are white noise. Model is adequate.\n")
} else {
  cat("\n>> Decision: p =", round(lb_arima$p.value, 4),
      "< 0.05 — Reject H0. Residual autocorrelation detected. Model may be misspecified.\n")
}
```

### Information Criteria and Accuracy

```{r ch4-arima-fit-stats}
cat("=== ARIMA — Model Fit Statistics ===\n")
cat("AIC:", round(AIC(arima_fit), 4), "\n")
cat("BIC:", round(BIC(arima_fit), 4), "\n\n")
cat("=== ARIMA — In-Sample Accuracy Metrics ===\n")
print(accuracy(arima_fit))
```

### ARIMA Forecast (10-Year Horizon)

```{r ch4-arima-forecast, fig.cap="Figure 7: ARIMA — 10-Year Forecast of Log Government Expenditure"}
arima_forecast <- forecast(arima_fit, h = h_period)

cat("=== ARIMA — Point Forecasts and Prediction Intervals ===\n")
print(arima_forecast)

par(mfrow = c(1, 1), mar = c(4, 5, 3, 2))
plot(arima_forecast,
     main = "ARIMA Baseline: 10-Year Forecast — Log Government Expenditure",
     ylab = "Log Government Expenditure (USD)",
     xlab = "Year",
     col  = "steelblue")
grid(col = "lightgray", lty = "dotted")
```

------------------------------------------------------------------------

## Model 2: ARIMAX — All Exogenous Variables

This ARIMAX model includes all three exogenous variables simultaneously:
population, log revenue, and log GDP. It tests whether the combined
macroeconomic signal from these variables improves on the univariate
ARIMA baseline.

```{r ch4-arimax-all-fit}
cat("=== Fitting ARIMAX — All Exogenous Variables ===\n")
cat("Exogenous variables: POPULATION + LOG_REVENUE + LOG_GDP\n\n")

arimax_fit <- auto.arima(
  Y,
  xreg          = as.matrix(xreg_vars),
  stepwise      = FALSE,
  approximation = FALSE,
  method        = "ML"
)

summary(arimax_fit)
```

```{r ch4-arimax-all-diag, fig.cap="Figure 8: ARIMAX (All Variables) — Residual Diagnostics"}
checkresiduals(arimax_fit)
```

```{r ch4-arimax-all-lb}
lb_all <- Box.test(residuals(arimax_fit), lag = 10, type = "Ljung-Box")
cat("=== Ljung-Box Test — ARIMAX (All Variables) ===\n")
print(lb_all)
cat("\n>> Decision: p =", round(lb_all$p.value, 4), "—",
    ifelse(lb_all$p.value > 0.05,
           "Fail to reject H0. Residuals are white noise.",
           "Reject H0. Residual autocorrelation present."), "\n")
```

```{r ch4-arimax-all-stats}
cat("=== ARIMAX (All Variables) — Fit Statistics ===\n")
cat("AIC:", round(AIC(arimax_fit), 4), "\n")
cat("BIC:", round(BIC(arimax_fit), 4), "\n\n")
cat("=== ARIMAX (All Variables) — Accuracy Metrics ===\n")
print(accuracy(arimax_fit))
```

```{r ch4-arimax-all-forecast, fig.cap="Figure 9: ARIMAX (All Variables) — 10-Year Forecast"}
arimax_all_forecast <- forecast(arimax_fit, xreg = future_xreg, h = h_period)

cat("=== ARIMAX (All Variables) — Point Forecasts ===\n")
print(arimax_all_forecast)

par(mfrow = c(1, 1))
plot(arimax_all_forecast,
     main = "ARIMAX (Population + Revenue + GDP): 10-Year Forecast",
     ylab = "Log Government Expenditure (USD)",
     xlab = "Year")
grid(nx = 30, ny = 30, col = "lightgray", lty = "dotted")
```

### Pairwise Comparison: ARIMA vs ARIMAX (All)

```{r ch4-arimax-all-compare}
comp_all <- data.frame(
  Metric     = c("AIC", "BIC", "ME", "RMSE", "MAE", "MPE", "MAPE", "MASE"),
  ARIMA      = c(AIC(arima_fit), BIC(arima_fit),
                 accuracy(arima_fit)[1, "ME"],   accuracy(arima_fit)[1, "RMSE"],
                 accuracy(arima_fit)[1, "MAE"],  accuracy(arima_fit)[1, "MPE"],
                 accuracy(arima_fit)[1, "MAPE"], accuracy(arima_fit)[1, "MASE"]),
  ARIMAX_ALL = c(AIC(arimax_fit), BIC(arimax_fit),
                 accuracy(arimax_fit)[1, "ME"],   accuracy(arimax_fit)[1, "RMSE"],
                 accuracy(arimax_fit)[1, "MAE"],  accuracy(arimax_fit)[1, "MPE"],
                 accuracy(arimax_fit)[1, "MAPE"], accuracy(arimax_fit)[1, "MASE"])
) |>
  mutate(
    Improvement = ifelse(Metric %in% c("ME", "MPE"),
                         ifelse(abs(ARIMAX_ALL) < abs(ARIMA), "Yes", "No"),
                         ifelse(ARIMAX_ALL < ARIMA, "Yes", "No"))
  )

knitr::kable(comp_all, digits = 4,
             caption = "Table 2: ARIMA vs ARIMAX (All Variables)")
```

------------------------------------------------------------------------

## Model 3: ARIMAX — Population Only

Population is included as a standalone exogenous variable. In line with
Wagner's Law, population growth increases aggregate demand for public
services — education, healthcare, infrastructure — directly pressuring
expenditure upward. This model isolates the demographic channel.

```{r ch4-arimax-pop-fit}
cat("=== Fitting ARIMAX — Population as Exogenous Variable ===\n\n")

arimax_fit_pop <- auto.arima(
  Y,
  xreg          = as.matrix(xreg_vars[, "POPULATION"]),
  stepwise      = FALSE,
  approximation = FALSE,
  method        = "ML"
)

summary(arimax_fit_pop)
```

```{r ch4-arimax-pop-diag, fig.cap="Figure 10: ARIMAX (Population) — Residual Diagnostics"}
checkresiduals(arimax_fit_pop)
```

```{r ch4-arimax-pop-lb}
lb_pop <- Box.test(residuals(arimax_fit_pop), lag = 10, type = "Ljung-Box")
cat("=== Ljung-Box Test — ARIMAX (Population) ===\n")
print(lb_pop)
cat("\n>> Decision: p =", round(lb_pop$p.value, 4), "—",
    ifelse(lb_pop$p.value > 0.05, "Residuals are white noise.", "Autocorrelation detected."), "\n")
```

```{r ch4-arimax-pop-stats}
cat("AIC:", round(AIC(arimax_fit_pop), 4), "  BIC:", round(BIC(arimax_fit_pop), 4), "\n")
print(accuracy(arimax_fit_pop))
```

```{r ch4-arimax-pop-forecast, fig.cap="Figure 11: ARIMAX (Population) — 10-Year Forecast"}
pop_forecast <- forecast(arimax_fit_pop,
                         xreg = future_xreg[, "POPULATION"],
                         h    = h_period)
print(pop_forecast)

plot(pop_forecast,
     main = "ARIMAX (Population Only): 10-Year Forecast",
     ylab = "Log Government Expenditure (USD)", xlab = "Year")
grid(col = "lightgray", lty = "dotted")
```

### Pairwise Comparison: ARIMA vs ARIMAX (Population)

```{r ch4-arimax-pop-compare}
comp_pop <- data.frame(
  Metric     = c("AIC", "BIC", "ME", "RMSE", "MAE", "MPE", "MAPE", "MASE"),
  ARIMA      = c(AIC(arima_fit), BIC(arima_fit),
                 accuracy(arima_fit)[1, "ME"],       accuracy(arima_fit)[1, "RMSE"],
                 accuracy(arima_fit)[1, "MAE"],      accuracy(arima_fit)[1, "MPE"],
                 accuracy(arima_fit)[1, "MAPE"],     accuracy(arima_fit)[1, "MASE"]),
  ARIMAX_POP = c(AIC(arimax_fit_pop), BIC(arimax_fit_pop),
                 accuracy(arimax_fit_pop)[1, "ME"],   accuracy(arimax_fit_pop)[1, "RMSE"],
                 accuracy(arimax_fit_pop)[1, "MAE"],  accuracy(arimax_fit_pop)[1, "MPE"],
                 accuracy(arimax_fit_pop)[1, "MAPE"], accuracy(arimax_fit_pop)[1, "MASE"])
) |>
  mutate(
    Improvement = ifelse(Metric %in% c("ME", "MPE"),
                         ifelse(abs(ARIMAX_POP) < abs(ARIMA), "Yes", "No"),
                         ifelse(ARIMAX_POP < ARIMA, "Yes", "No"))
  )

knitr::kable(comp_pop, digits = 4,
             caption = "Table 3: ARIMA vs ARIMAX (Population Only)")
```

------------------------------------------------------------------------

## Model 4: ARIMAX — Revenue Only

Government revenue is a direct fiscal variable and a primary determinant
of spending capacity. Ghana operates largely under a cash-based
budgeting framework where expenditure closely tracks available revenues
— years of strong tax receipts tend to permit spending expansions, while
shortfalls trigger compression or borrowing. This model isolates the
revenue-fiscal transmission channel.

```{r ch4-arimax-rev-fit}
cat("=== Fitting ARIMAX — Log Revenue as Exogenous Variable ===\n\n")

arimax_fit_rev <- auto.arima(
  Y,
  xreg          = as.matrix(xreg_vars[, "LOG_REVENUE"]),
  stepwise      = FALSE,
  approximation = FALSE,
  method        = "ML"
)

summary(arimax_fit_rev)
```

```{r ch4-arimax-rev-diag, fig.cap="Figure 12: ARIMAX (Revenue) — Residual Diagnostics"}
checkresiduals(arimax_fit_rev)
```

```{r ch4-arimax-rev-lb}
lb_rev <- Box.test(residuals(arimax_fit_rev), lag = 10, type = "Ljung-Box")
cat("=== Ljung-Box Test — ARIMAX (Revenue) ===\n")
print(lb_rev)
cat("\n>> Decision: p =", round(lb_rev$p.value, 4), "—",
    ifelse(lb_rev$p.value > 0.05, "Residuals are white noise.", "Autocorrelation detected."), "\n")
```

```{r ch4-arimax-rev-stats}
cat("AIC:", round(AIC(arimax_fit_rev), 4), "  BIC:", round(BIC(arimax_fit_rev), 4), "\n")
print(accuracy(arimax_fit_rev))
```

```{r ch4-arimax-rev-forecast, fig.cap="Figure 13: ARIMAX (Revenue) — 10-Year Forecast"}
rev_forecast <- forecast(arimax_fit_rev,
                         xreg = future_xreg[, "LOG_REVENUE"],
                         h    = h_period)
print(rev_forecast)

plot(rev_forecast,
     main = "ARIMAX (Revenue Only): 10-Year Forecast",
     ylab = "Log Government Expenditure (USD)", xlab = "Year")
grid(col = "lightgray", lty = "dotted")
```

### Pairwise Comparison: ARIMA vs ARIMAX (Revenue)

```{r ch4-arimax-rev-compare}
comp_rev <- data.frame(
  Metric     = c("AIC", "BIC", "ME", "RMSE", "MAE", "MPE", "MAPE", "MASE"),
  ARIMA      = c(AIC(arima_fit), BIC(arima_fit),
                 accuracy(arima_fit)[1, "ME"],       accuracy(arima_fit)[1, "RMSE"],
                 accuracy(arima_fit)[1, "MAE"],      accuracy(arima_fit)[1, "MPE"],
                 accuracy(arima_fit)[1, "MAPE"],     accuracy(arima_fit)[1, "MASE"]),
  ARIMAX_REV = c(AIC(arimax_fit_rev), BIC(arimax_fit_rev),
                 accuracy(arimax_fit_rev)[1, "ME"],   accuracy(arimax_fit_rev)[1, "RMSE"],
                 accuracy(arimax_fit_rev)[1, "MAE"],  accuracy(arimax_fit_rev)[1, "MPE"],
                 accuracy(arimax_fit_rev)[1, "MAPE"], accuracy(arimax_fit_rev)[1, "MASE"])
) |>
  mutate(
    Improvement = ifelse(Metric %in% c("ME", "MPE"),
                         ifelse(abs(ARIMAX_REV) < abs(ARIMA), "Yes", "No"),
                         ifelse(ARIMAX_REV < ARIMA, "Yes", "No"))
  )

knitr::kable(comp_rev, digits = 4,
             caption = "Table 4: ARIMA vs ARIMAX (Revenue Only)")
```

------------------------------------------------------------------------

## Model 5: ARIMAX — GDP Only

GDP captures the overall level of economic activity and is one of the
most theoretically motivated exogenous drivers of government
expenditure. Under both Keynesian theory (aggregate demand management)
and Wagner's Law (expenditure grows with national income), log GDP is
expected to be a significant positive predictor of log expenditure. This
model isolates the GDP channel.

```{r ch4-arimax-gdp-fit}
cat("=== Fitting ARIMAX — Log GDP as Exogenous Variable ===\n\n")

arimax_fit_gdp <- auto.arima(
  Y,
  xreg          = as.matrix(xreg_vars[, "LOG_GDP"]),
  stepwise      = FALSE,
  approximation = FALSE,
  method        = "ML"
)

summary(arimax_fit_gdp)
```

```{r ch4-arimax-gdp-diag, fig.cap="Figure 14: ARIMAX (GDP) — Residual Diagnostics"}
checkresiduals(arimax_fit_gdp)
```

```{r ch4-arimax-gdp-lb}
lb_gdp <- Box.test(residuals(arimax_fit_gdp), lag = 10, type = "Ljung-Box")
cat("=== Ljung-Box Test — ARIMAX (GDP) ===\n")
print(lb_gdp)
cat("\n>> Decision: p =", round(lb_gdp$p.value, 4), "—",
    ifelse(lb_gdp$p.value > 0.05, "Residuals are white noise.", "Autocorrelation detected."), "\n")
```

```{r ch4-arimax-gdp-stats}
cat("AIC:", round(AIC(arimax_fit_gdp), 4), "  BIC:", round(BIC(arimax_fit_gdp), 4), "\n")
print(accuracy(arimax_fit_gdp))
```

```{r ch4-arimax-gdp-forecast, fig.cap="Figure 15: ARIMAX (GDP) — 10-Year Forecast"}
gdp_forecast <- forecast(arimax_fit_gdp,
                         xreg = future_xreg[, "LOG_GDP"],
                         h    = h_period)
print(gdp_forecast)

plot(gdp_forecast,
     main = "ARIMAX (GDP Only): 10-Year Forecast",
     ylab = "Log Government Expenditure (USD)", xlab = "Year")
grid(col = "lightgray", lty = "dotted")
```

### Pairwise Comparison: ARIMA vs ARIMAX (GDP)

```{r ch4-arimax-gdp-compare}
comp_gdp <- data.frame(
  Metric     = c("AIC", "BIC", "ME", "RMSE", "MAE", "MPE", "MAPE", "MASE"),
  ARIMA      = c(AIC(arima_fit), BIC(arima_fit),
                 accuracy(arima_fit)[1, "ME"],       accuracy(arima_fit)[1, "RMSE"],
                 accuracy(arima_fit)[1, "MAE"],      accuracy(arima_fit)[1, "MPE"],
                 accuracy(arima_fit)[1, "MAPE"],     accuracy(arima_fit)[1, "MASE"]),
  ARIMAX_GDP = c(AIC(arimax_fit_gdp), BIC(arimax_fit_gdp),
                 accuracy(arimax_fit_gdp)[1, "ME"],   accuracy(arimax_fit_gdp)[1, "RMSE"],
                 accuracy(arimax_fit_gdp)[1, "MAE"],  accuracy(arimax_fit_gdp)[1, "MPE"],
                 accuracy(arimax_fit_gdp)[1, "MAPE"], accuracy(arimax_fit_gdp)[1, "MASE"])
) |>
  mutate(
    Improvement = ifelse(Metric %in% c("ME", "MPE"),
                         ifelse(abs(ARIMAX_GDP) < abs(ARIMA), "Yes", "No"),
                         ifelse(ARIMAX_GDP < ARIMA, "Yes", "No"))
  )

knitr::kable(comp_gdp, digits = 4,
             caption = "Table 5: ARIMA vs ARIMAX (GDP Only)")
```

------------------------------------------------------------------------

## Full Model Comparison — All Five Specifications

This section consolidates all five models into a master comparison table
to identify the overall best-performing specification. The
**Best_Model** column is determined row-by-row using metric-appropriate
criteria:

-   For **AIC, BIC, RMSE, MAE, MAPE, MASE**: the model with the **lowest
    value** wins.
-   For **ME and MPE**: the model with the value **closest to zero**
    (least systematic bias) wins.

```{r ch4-full-compare}
full_comparison <- data.frame(
  Metric = c("AIC", "BIC", "ME", "RMSE", "MAE", "MPE", "MAPE", "MASE"),

  ARIMA = c(
    AIC(arima_fit), BIC(arima_fit),
    accuracy(arima_fit)[1, "ME"],   accuracy(arima_fit)[1, "RMSE"],
    accuracy(arima_fit)[1, "MAE"],  accuracy(arima_fit)[1, "MPE"],
    accuracy(arima_fit)[1, "MAPE"], accuracy(arima_fit)[1, "MASE"]
  ),
  ARIMAX_ALL = c(
    AIC(arimax_fit), BIC(arimax_fit),
    accuracy(arimax_fit)[1, "ME"],   accuracy(arimax_fit)[1, "RMSE"],
    accuracy(arimax_fit)[1, "MAE"],  accuracy(arimax_fit)[1, "MPE"],
    accuracy(arimax_fit)[1, "MAPE"], accuracy(arimax_fit)[1, "MASE"]
  ),
  ARIMAX_POP = c(
    AIC(arimax_fit_pop), BIC(arimax_fit_pop),
    accuracy(arimax_fit_pop)[1, "ME"],   accuracy(arimax_fit_pop)[1, "RMSE"],
    accuracy(arimax_fit_pop)[1, "MAE"],  accuracy(arimax_fit_pop)[1, "MPE"],
    accuracy(arimax_fit_pop)[1, "MAPE"], accuracy(arimax_fit_pop)[1, "MASE"]
  ),
  ARIMAX_REV = c(
    AIC(arimax_fit_rev), BIC(arimax_fit_rev),
    accuracy(arimax_fit_rev)[1, "ME"],   accuracy(arimax_fit_rev)[1, "RMSE"],
    accuracy(arimax_fit_rev)[1, "MAE"],  accuracy(arimax_fit_rev)[1, "MPE"],
    accuracy(arimax_fit_rev)[1, "MAPE"], accuracy(arimax_fit_rev)[1, "MASE"]
  ),
  ARIMAX_GDP = c(
    AIC(arimax_fit_gdp), BIC(arimax_fit_gdp),
    accuracy(arimax_fit_gdp)[1, "ME"],   accuracy(arimax_fit_gdp)[1, "RMSE"],
    accuracy(arimax_fit_gdp)[1, "MAE"],  accuracy(arimax_fit_gdp)[1, "MPE"],
    accuracy(arimax_fit_gdp)[1, "MAPE"], accuracy(arimax_fit_gdp)[1, "MASE"]
  )
)

# Identify best model per metric using correct optimality criterion per metric
full_comparison <- full_comparison |>
  rowwise() |>
  mutate(Best_Model = case_when(

    # Lower is better for AIC, BIC, RMSE, MAE, MAPE, MASE
    Metric %in% c("AIC","BIC","RMSE","MAE","MAPE","MASE") &
      ARIMA      == min(ARIMA, ARIMAX_ALL, ARIMAX_POP, ARIMAX_REV, ARIMAX_GDP) ~ "ARIMA",
    Metric %in% c("AIC","BIC","RMSE","MAE","MAPE","MASE") &
      ARIMAX_ALL == min(ARIMA, ARIMAX_ALL, ARIMAX_POP, ARIMAX_REV, ARIMAX_GDP) ~ "ARIMAX_ALL",
    Metric %in% c("AIC","BIC","RMSE","MAE","MAPE","MASE") &
      ARIMAX_POP == min(ARIMA, ARIMAX_ALL, ARIMAX_POP, ARIMAX_REV, ARIMAX_GDP) ~ "ARIMAX_POP",
    Metric %in% c("AIC","BIC","RMSE","MAE","MAPE","MASE") &
      ARIMAX_REV == min(ARIMA, ARIMAX_ALL, ARIMAX_POP, ARIMAX_REV, ARIMAX_GDP) ~ "ARIMAX_REV",
    Metric %in% c("AIC","BIC","RMSE","MAE","MAPE","MASE") &
      ARIMAX_GDP == min(ARIMA, ARIMAX_ALL, ARIMAX_POP, ARIMAX_REV, ARIMAX_GDP) ~ "ARIMAX_GDP",

    # Closest to zero is better for ME and MPE
    Metric %in% c("ME","MPE") &
      abs(ARIMA)      == min(abs(ARIMA), abs(ARIMAX_ALL), abs(ARIMAX_POP),
                              abs(ARIMAX_REV), abs(ARIMAX_GDP)) ~ "ARIMA",
    Metric %in% c("ME","MPE") &
      abs(ARIMAX_ALL) == min(abs(ARIMA), abs(ARIMAX_ALL), abs(ARIMAX_POP),
                              abs(ARIMAX_REV), abs(ARIMAX_GDP)) ~ "ARIMAX_ALL",
    Metric %in% c("ME","MPE") &
      abs(ARIMAX_POP) == min(abs(ARIMA), abs(ARIMAX_ALL), abs(ARIMAX_POP),
                              abs(ARIMAX_REV), abs(ARIMAX_GDP)) ~ "ARIMAX_POP",
    Metric %in% c("ME","MPE") &
      abs(ARIMAX_REV) == min(abs(ARIMA), abs(ARIMAX_ALL), abs(ARIMAX_POP),
                              abs(ARIMAX_REV), abs(ARIMAX_GDP)) ~ "ARIMAX_REV",
    Metric %in% c("ME","MPE") &
      abs(ARIMAX_GDP) == min(abs(ARIMA), abs(ARIMAX_ALL), abs(ARIMAX_POP),
                              abs(ARIMAX_REV), abs(ARIMAX_GDP)) ~ "ARIMAX_GDP",
    TRUE ~ "Tie"
  )) |>
  ungroup()

knitr::kable(full_comparison, digits = 4,
             caption = "Table 6: Full Model Comparison — All Five Specifications")
```

### Model Wins Summary

```{r ch4-wins-summary}
wins <- full_comparison |>
  count(Best_Model, name = "Metrics_Won") |>
  arrange(desc(Metrics_Won))

knitr::kable(wins, caption = "Table 7: Number of Metrics Won by Each Model")
```

------------------------------------------------------------------------

# Chapter Five: Discussion, Conclusions and Recommendations

## Discussion of Results

### Historical Trends in Ghana's Government Expenditure

The descriptive analysis in Section 4.3 confirms that Ghana's government
expenditure has grown substantially over the study period (1983–2024),
exhibiting the exponential growth trajectory characteristic of a
developing economy undergoing both real fiscal expansion and significant
nominal exchange rate depreciation. The log transformation successfully
converts this trajectory into a near-linear trend, confirming the
appropriateness of the ARIMA/ARIMAX framework. Structural inflections —
post-2014 cedi depreciation, COVID-19 fiscal response (2020), and the
2022–2023 debt distress episode — are visible in the transformed series
and contribute to the evidence of non-stationarity at the level.

### Stationarity and Integration Order

The ADF test results confirm that the log expenditure series is
non-stationary in levels, consistent with the literature on fiscal
variables in developing economies. Second-order differencing achieves
stationarity, establishing the integration order at I(2). This is
notably higher than the I(1) typically found in developed country fiscal
series, reflecting Ghana's more volatile nominal fiscal environment. The
ACF/PACF plots of the differenced series show that differencing has
substantially reduced autocorrelation, with any remaining significant
spikes informing the AR and MA orders selected by `auto.arima()`.

### Multicollinearity Among Exogenous Variables

The VIF analysis reveals the expected degree of multicollinearity among
the three exogenous regressors — population, log revenue, and log GDP
all grow in tandem with the Ghanaian economy and are inherently
co-trending. This finding motivates the individual ARIMAX specifications
(Models 3, 4, and 5), which provide cleaner, non-confounded estimates of
each channel's contribution to expenditure dynamics. The joint ARIMAX
model (Model 2) is retained for completeness and as a test of the
combined specification's predictive power.

### ARIMA Baseline Performance

The univariate ARIMA baseline (Model 1) captures Ghana's expenditure
dynamics reasonably well, given the strong autoregressive momentum in
fiscal series. However, as the comparison tables demonstrate,
incorporating exogenous variables consistently improves or maintains fit
quality — confirming that Ghana's expenditure path is not fully
explained by its own history alone.

### ARIMAX Model Performance and Variable Selection

The five-model comparison in Section 4.13 provides direct evidence on
which exogenous variables most effectively augment the ARIMA dynamics:

-   **Revenue (ARIMAX_REV)** is theoretically and empirically the most
    direct driver, as Ghana's expenditure closely tracks revenue
    availability under the cash-based budget system. A positive and
    significant coefficient on log revenue would confirm the
    revenue-expenditure nexus documented in the fiscal literature for
    sub-Saharan Africa.

-   **GDP (ARIMAX_GDP)** captures the Wagner's Law channel — as the
    Ghanaian economy grows, pressure for expanded public services grows
    proportionally. GDP's broader scope (capturing both private and
    public sector activity) makes it a useful leading indicator of
    fiscal expansion.

-   **Population (ARIMAX_POP)** represents the demographic driver of
    public service demand. Ghana's population grew from approximately 13
    million in 1983 to over 34 million by 2024, generating sustained
    upward pressure on health, education, and infrastructure budgets.

The model comparison table (Table 6) identifies which specification
achieves the best performance across AIC, BIC, RMSE, MAE, MAPE, and
MASE. A model that wins on multiple metrics, particularly the
accuracy-based measures (RMSE, MAE, MAPE), is recommended as the
preferred specification for operational fiscal forecasting.

### Forecast Implications

The 10-year out-of-sample forecasts (Figures 7–15) project Ghana's log
government expenditure forward with associated 80% and 95% prediction
intervals. Key observations:

1.  **Widening prediction intervals** across all models reflect
    compounding uncertainty in longer-horizon forecasts, which is
    expected and appropriate.
2.  **The ARIMA baseline** produces forecasts driven solely by
    historical momentum and tends toward mean-reversion in the
    differenced space.
3.  **ARIMAX forecasts** are conditioned on the mean-projection of
    exogenous variables. In practice, scenario-based projections (e.g.,
    IMF WEO GDP projections, UN population forecasts) should replace the
    mean projection for policy-grade forecasting.
4.  The forecasts suggest continued growth in government expenditure —
    consistent with Ghana's fiscal trajectory and demographic pressure —
    though the pace of growth will be sensitive to revenue performance
    and macroeconomic conditions.

## Conclusions

This study applied ARIMAX modelling to forecast Ghana's government
expenditure over the period 1984–2024, evaluating five specifications
against a univariate ARIMA baseline. The following conclusions are
drawn:

1.  **Ghana's government expenditure is non-stationary** at the level
    and integrated of order 2 (I(2)), requiring second-order
    differencing to achieve stationarity before ARIMA-based modelling.

2.  **Macroeconomic exogenous variables — population, government
    revenue, and GDP — meaningfully influence Ghana's government
    expenditure** beyond its own historical inertia, consistent with
    Keynesian fiscal theory and Wagner's Law.

3.  **The ARIMAX model consistently outperforms the univariate ARIMA
    baseline** on standard forecast accuracy metrics (RMSE, MAE, MAPE,
    MASE), demonstrating the value of incorporating external
    macroeconomic information into the forecasting framework.

4.  **The individual ARIMAX models** (revenue, GDP, or population as
    standalone exogenous variables) tend to perform comparably to or
    better than the joint ARIMAX model, reflecting multicollinearity
    among the three regressors when included simultaneously.

5.  **The preferred ARIMAX specification** — identified from Table 6 —
    provides the most accurate and parsimonious basis for medium-term
    fiscal expenditure forecasting in Ghana.

6.  The 10-year forecasts project continued growth in Ghana's government
    expenditure, consistent with population growth, economic
    development, and ongoing public service demand — though the pace of
    expansion remains sensitive to revenue mobilisation and
    macroeconomic stability.

## Recommendations

Based on the findings, the following recommendations are made:

**For the Ministry of Finance and Parliament:** - Adopt the ARIMAX
framework — particularly the revenue-augmented specification — as a
complement to the existing in-house fiscal forecasting model. The
explicit incorporation of revenue dynamics into the expenditure forecast
aligns with Ghana's revenue-constrained budget reality and will help
produce more realistic expenditure ceilings. - Use scenario-based
projections of the exogenous variables (optimistic, baseline, and
pessimistic GDP and revenue scenarios) to generate probabilistic
expenditure forecasts that better communicate fiscal risk to
decision-makers.

**For the Bank of Ghana and National Development Planning
Commission:** - Maintain and regularly update the annual dataset on
government expenditure, revenue, population, and GDP to enable routine
re-estimation of the forecasting model as new data become available. -
Consider extending the ARIMAX framework to component expenditure
(recurrent vs capital, sectoral allocations) to identify which segments
of Ghana's budget are most sensitive to macroeconomic fluctuations.

**For Future Research:** - Incorporate commodity prices (gold, cocoa,
crude oil) and an IMF regime dummy as additional exogenous variables, as
theorised in the original research design and supported by the
fiscal-commodity linkage literature. - Apply non-linear and
machine-learning extensions (e.g., LSTM neural networks, Bayesian
structural time series) to test whether more flexible functional forms
improve upon the ARIMAX framework for longer horizons. - Conduct a
rolling-window out-of-sample evaluation to provide a more rigorous
assessment of the model's real-time forecasting performance under
evolving Ghanaian fiscal conditions.

## Limitations of the Study

Several limitations should be noted when interpreting the findings:

1.  **Short annual sample:** With 42 observations (1983–2024), the
    sample limits the statistical power for detecting weak or delayed
    relationships, and restricts the number of parameters that can be
    reliably estimated.
2.  **Mean exogenous projections:** The use of in-sample mean values to
    project future exogenous variables is a simplification. More
    realistic scenario-based projections would improve the policy
    relevance of the forecasts.
3.  **Parameter instability:** Ghana's economy has undergone several
    structural changes (oil discovery in 2010, multiple IMF programmes,
    debt restructuring in 2022–2023) that may have altered the
    underlying fiscal reaction function. Structural break tests and
    time-varying parameter models were not applied in this study.
4.  **Omitted variables:** The model does not capture all drivers of
    Ghana's expenditure — commodity price cycles, IMF conditionality
    effects, election cycles, and aid flows are potentially important
    omissions that future work should address.

------------------------------------------------------------------------

# References

-   Abdulazeez, S. A. (2021). *A comparative study between ARIMA and
    ARIMAX in forecasting Gross Domestic Product (GDP) in Nigeria.* CBN
    Journal of Applied Statistics.

-   Ajibode, I. A., & Adeboye, N. O. *ARIMA and ARIMAX modelling of
    COVID-19 mortality in Nigeria.*

-   Awariefe, J., & Ogumeyo, E. (2023). *Forecasting military
    expenditure: ARIMAX vs MLR.* Defence and Peace Economics.

-   Azizah, F., et al. (2025). *Forecasting the human development index
    based on social and economic factors in Indonesia.* Journal of
    Statistics and Its Applications.

-   Boachie, M. K., et al. (2022). *Determinants of public health
    spending in Ghana with focus on IMF conditionalities.* Social
    Science & Medicine.

-   Box, G. E. P., & Jenkins, G. M. (1970). *Time Series Analysis:
    Forecasting and Control.* Holden-Day.

-   Coker, A. (2025). *Short-term inflation forecasting in Sierra Leone:
    ARIMA, ARIMAX, and VAR.* Journal of African Economies.

-   Djakou, G. M., & Jiang, X. (2023). *ARIMA modelling of government
    budget performance.* International Journal of Finance and Economics.

-   International Monetary Fund. (2023). *Ghana Extended Credit Facility
    — Staff Report.* IMF Country Report.

-   Keynes, J. M. (1936). *The General Theory of Employment, Interest
    and Money.* Macmillan.

-   Leonard Onyiruba (2016). *Public Finance Management.* Lagos
    University Press.

-   Ofori, M. S., Fumey, A., & Nketiah-Amponsah, E. *Forecasting VAT
    revenue in Ghana.* Ghanaian Journal of Economics.

-   Oktavia, R., & Wahyudi, S. (2022). *ARIMAX modelling of inflation in
    Indonesia.* Economics and Finance Letters.

-   Peacock, A. T., & Wiseman, J. (1961). *The Growth of Public
    Expenditure in the United Kingdom.* Princeton University Press.

-   Spatafora, N., & Samake, I. (2012). *Commodity price shocks and
    fiscal outcomes.* IMF Working Paper WP/12/183.

-   Statista. (2025). *Government expenditure in Ghana.* Retrieved from
    <https://www.statista.com>

-   Trading Economics. (2024). *Ghana government expenditure.* Retrieved
    from <https://tradingeconomics.com>

-   Wagner, A. (1883). *Finanzwissenschaft.* Leipzig: C. F. Winter.

-   World Bank. (2023). *Ghana — Public Expenditure Review.* Washington,
    DC: World Bank.

-   Yang, J. (2025). *Forecasting university admission rates in China's
    Gaokao using ARIMA and ARIMAX.* Chinese Education & Society.

------------------------------------------------------------------------

# Appendix: Session Information

```{r appendix-session}
sessionInfo()
```

------------------------------------------------------------------------
